---
title: "Analysis between Diabetes and BMI using Logistic Model"
author: "Jingyao Wang"
date: "December 2020"
output: pdf_document
code_folding: "hide"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\section*{Abstract}
Studies have shown that diabetes is one of the most serious diseases affecting people's lives. We used the Pima Indians Diabetes Database to observe various factors that affect diabetes. Among them, we found that obesity is closely related to diabetes. A higher BMI indicates a higher chance of getting diabetes.

\section*{Introduction}
Diabetes is currently one of the most common metabolic diseases, and its main feature is hyperglycemia. If blood sugar levels cannot be controlled, long-term high blood sugar can cause serious damage to organs and nerves. For example, blindness and kidney failure are common complications of diabetes. According to the research of WHO, in 2014, over 8.5\% people who are older than 18 had diabetes. Moreover, 1.6 million people died due to high blood sugar in 2016. WHO estimates that one of the major causes of death in 2016 is diabetes. Between 2000 and 2016, the number of premature deaths due to diabetes showed an upward trend. Due to the number of people suffering from diabetes has continued to rise in recent years, they have also paid more and more attention to it. For this reason, many preventive measures have been summarized, and the hospital will also check related factors carefully during the physical examination. Since diabetes is not an infectious disease, people can control or prevent it by adjusting their living habits. Fitness, healthy eating and avoiding staying up late are highly recommended for maintaining body in a healthy status.

Type 2 Diabetes is highly related to obesity. Research demonstrates that 85\% of people with type 2 diabetes are overweight. Body mass index (BMI) is an index that measure body fat. By calculating BMI, we can intuitively know whether a person is obese or not. We define obesity as a BMI over 25.0.


```{r, include=FALSE}
library(tidyverse)
library(broom)
library(arm)
library(gridExtra)
library(kableExtra)
```

```{r, include=FALSE}
data = read.csv("diabetes.csv",header = TRUE)

                                             
summary(data)
```

```{r, include=FALSE}
data <-
  data %>% 
  mutate(BMI_num = case_when(
    BMI >= 25 ~1,
    BMI < 25 ~ 0))


propensity_score <- glm(BMI_num ~ Pregnancies + Glucose + Insulin + SkinThickness +DiabetesPedigreeFunction + Age, 
                        family = binomial,
                        data = data)

data <- 
  augment(propensity_score, 
          data = data,
          type.predict = "response") %>% 
  dplyr::select(-.resid, -.std.resid, -.hat, -.sigma, -.cooksd) 

data <- 
  data %>% 
  arrange(.fitted, BMI)

data$treated <- 
  if_else(data$BMI == 0, 0, 1)

data$treated <- 
  as.integer(data$treated)

matches <- arm::matching(z = data$treated, 
                         score = data$.fitted)

data <- cbind(data, matches)

data_matched <- 
  data %>% 
  filter(match.ind != 0) %>% 
  dplyr::select(-match.ind, -pairs, -treated)

head(data_matched)
```

\section*{Data}
We used a set of data that is released by the National Institute of Diabetes and Digestive and Kidney Diseases. It contains eight diagnostic measurements (variables) related to diabetes. Through these variables, we can judge the connection with the diagnosis result. 769 observations were collected with different factors that related to diabetes. Since obesity is one of the most common problems at present, we want to study the relationship between BMI and other variables first.
Propensity matching score was applied for matching the treated and controlled observations. By using this method, we can minimize the bias between the treatment group and our final interest outcome. The data of BMI were separated into two different levels,smaller than 25.0 and lager or equal to 25.0. We created a new binary variable named \textit{BMI\_num} in order to set up the logstic model. \textit{Pregnancies}. \textit{Glucose}, \textit{BloodPressure}, \textit{SkinThickness}, \textit{Insulin}, \textit{DiabetesPedigreeFunction} and \textit{Age} are all considered as variables of this model.

```{r, echo=FALSE}
dt1 <- c("Pregnancies", "Glucose" , "BloodPressure" , "SkinThickness" , "Insulin", "DiabetesPedigreeFunction", "Age")

dt2 <- c("times of pregnancy" , "concentration of plasma glucose in an oral glucose tolerance test" , "blood pressure", "thickness of triceps skin fold" , "serum insulin in 2 hrs", "a function that score the probability of having diabetes based on family history", "The age of observations")


res <- data.frame(
  Variables=dt1, Meanings=dt2
  
  )

knitr::kable(res,
             caption = "Meaning of Variables") %>%
  kable_styling(latex_options = "hold_position")
```

\section*{Model}
Now we need to build a logistic model to predict the result of \textit{Outcome}, which is whether an observation is having diabetes or not. By the logsitic model, we can easily find out how these 7 variables affects the outcome of our prediction.

\begin{equation}
\begin{split}
log(\frac{p}{1-p}) =& \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \beta_4 x_4 + \beta_5 x_5 + \beta_6 x_6 + \beta_7 x_7 + \beta_8 x_8     
\end{split}
\end{equation}

\begin{equation}
\begin{split}
\frac{p}{1-p} = e^{\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \beta_4 x_4 + \beta_5 x_5 + \beta_6 x_6 + \beta_7 x_7 + \beta_8 x_8}
\end{split}
\end{equation}




```{r, echo = FALSE, message=FALSE, include=FALSE}
propensity_score_regression <- 
  glm(Outcome ~ Pregnancies + Glucose + Insulin + SkinThickness +DiabetesPedigreeFunction + Age + BMI,
      data = data_matched)

huxtable::huxreg(propensity_score_regression)
```

```{r, include=FALSE}
summary(propensity_score_regression)
```

\section*{Result}
We summarise the result of \textit{Outcome} and \textit{BMI} by ploting these two variables with R.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
Diabetes <- data %>%
  group_by(Outcome) %>%
  summarise(counts = n())

BMIsum <-data %>% 
  group_by(BMI) %>%
  summarise(counts = n())

plot1 <- ggplot(Diabetes,aes(x=Outcome, y=counts))+
  geom_bar(fill = "grey", stat = "identity") +
  geom_text(aes(label = counts)) + 
  labs(title = "Barplot for Diabetes",
       subtitle = "Figure_1") +
  theme() +
  coord_flip()

plot2 <- ggplot(BMIsum,aes(x=BMI, y=counts))+
  geom_bar(fill = "grey", stat = "identity") +
  geom_text(aes(label = counts)) + 
  labs(title = "Plot for BMI",
       subtitle = "Figure_2") +
  theme() +
  coord_flip()

grid.arrange(plot1, plot2, nrow=2)
```


\subsection*{Result of Logistic Model}
```{r, echo=FALSE}
dt1 <- c("(Intercept)","Pregnancies", "Glucose" , "BloodPressure" , "SkinThickness" , "Insulin", "DiabetesPedigreeFunction", "Age", "BMI")

dt2 <- c("-0.192866" , "0.058395" , "0.001185", "0.005582" , "0.005710", "0.005582", "-0.402789", "0.005358", "0.002006")


res2 <- data.frame(
  Variables=dt1, Coefficients=dt2
  
  )

knitr::kable(res2,
             caption = "Coefficients of Variables") %>%
  kable_styling(latex_options = "hold_position")
```

\textbf{Interpretation of $\beta_k \left(k=0,1,...,8\right)$}:

$\beta_1$:
As times of pregnancy increasing, log odds of having diabetes will increase by 0.058395

$\beta_2$:
As the concentration of plasma glucose in an oral glucose tolerance test increasing, log odds of having diabetes will increase by 0.001185

$\beta_3$:
As the blood pressure increasing, log odds of having diabetes will increase by 0.005582

$\beta_4$:
As thickness of triceps skin fold, log odds of having diabetes will increase by 0.005710

$\beta_5$:
As the Insulin increasing, log odds of having diabetes will increase by 0.005582

$\beta_6$:
As the score of Diabetes Pedigree Function increasing, log odds of having diabetes will decrease by 0.402789

$\beta_7$:
As the age increasing, log odds of having diabetes will increase by 0.005358

$\beta_8$:
As the BMI increasing, log odds of having diabetes will increase by 0.002006

\textbf{Interpretation of Logsitic Model}:

Since the type of \textit{Outcome} is binary, we choose logisit model to make the assumption. The coefficients of the variables indicates that the probabilty of one having diabetes show an upward trend as the index of these variables increasing, except DiabetesPedigreeFunction.

\section*{Discussion}
The coefficient of BMI\_num is -0.063085 which implies that a person with lower BMI is less likelihood have diabetes. It is essential to keep body weight within a normal range. According to Table 2 above, we noticed that coefficients of  \textit{Pregnancies}, \textit{BloodPressure}, \textit{SkinThickness}, \textit{Insulin} and \textit{Age} are positive, which means that the rise of these variables shows a positive correlation with diabetes. People who are experiencing a higher insulin level often  associated with Hyperinsulinemia, which is related to diabetes.

People who experienced more times of pregnancies are likely to have a higher chance of getting diabetes than others. Move on to the concentration of plasma glucose in an oral glucose tolerance test, a higher index indicates a higher probability of having diabetes. And as the age increases, the risk of diabetes is also increasing.

Furthermore, through the analysis results, we can roughly understand how to reduce the risk of diabetes. First, control our body weight, which requires a healthy diet and exercise. Second, control the intake of sugar. Last but not least, since the risk of diabetes will increase with aging, regular physical examinations are also very important.

\subsection*{Weakness}
We are very satisfied that the result of $\beta$ is in line with our expectation. Besides, all the variables are independent which makes the bias smaller. However, there are several weakness of the model. The sample size of this dataset is not that large. It is hard for us to make a percise conclusion. Also, even though all of the variables are highly related to our interest outcome, it would be better to include the gender, Country of Citizenship or A1C result in order to compare the difference among countries, female or male.

\subsection*{Next step}
Next we want to use a more complex model such as Bayer’s model to make predictions in order to get more accurate results. Also, since obesity is not the only reason for diabetes, we would like to find out the connections between the disease and other factors.

\section*{Reference}
Pima Indians Diabetes Database
UCI Learning - https://www.kaggle.com/uciml/pima-indians-diabetes-database

Diabetes
https://www.who.int/news-room/fact-sheets/detail/diabetes

Body Mass Index (bmi) Calculator
https://www.diabetes.ca/managing-my-diabetes/tools---resources/body-mass-index-(bmi)-calculator

Hyperinsulinemia: Is It Diabetes?
https://www.mayoclinic.org/diseases-conditions/type-2-diabetes/expert-answers/hyperinsulinemia/faq-20058488

The Uk Is the Fattest Country in Europe. The Number Of Obese Adults Is Forecast To Rise By 73% Over the Next 20 Years from To 26 Million People, Resulting in More Than a Million Extra Cases Of Type 2 Diabetes, Heart Disease and Cancer.
15th January 2019 By Editor- Editor - https://www.diabetes.co.uk/diabetes-and-obesity.html

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open
  Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686

David Robinson, Alex Hayes and Simon Couch (2020). broom: Convert Statistical
  Objects into Tidy Tibbles. R package version 0.7.3.
  https://CRAN.R-project.org/package=broom

Andrew Gelman and Yu-Sung Su (2020). arm: Data Analysis Using Regression and
  Multilevel/Hierarchical Models. R package version 1.11-2.
  https://CRAN.R-project.org/package=arm
  
Baptiste Auguie (2017). gridExtra: Miscellaneous Functions for "Grid" Graphics. R
  package version 2.3. https://CRAN.R-project.org/package=gridExtra

Hao Zhu (2020). kableExtra: Construct Complex Table with 'kable' and Pipe Syntax.
  R package version 1.2.1. https://CRAN.R-project.org/package=kableExtra
  

\section*{Appendix}
Code and data supporting this anaylysis is available at: https://github.com/Wang-Lucy107/Analysis-of-Diabetes
